---
title: "Disease_of_pulmonary_TB_and_Pneumonia"
output:
  html_document:
    df_print: paged
---

```{r}
# set path
setwd("C:/Users/DELL/Dropbox/My PC (DESKTOP-6M0063Q)/Desktop/Stat_CU/Time Series/Project")
library("readxl")
data<- read_excel("Predict Pulmonary_tb(P'Pang).xlsx",sheet = "Time_series_data")
```

# plot Time Series Pneumonia
```{r}
ts_pneu <- ts(data$`Sum of Pneumonia`,start = c(2009,1),frequency = 12)
plot(ts_pneu)

```
Split data training and testing (test on 2 year)
```{r}
train_pneu <- ts(ts_pneu[c(1:108)],start = c(2009,1),frequency = 12)
test_pneu <- ts(ts_pneu[c(109:132)],start = c(2018,1),frequency = 12)
plot(train_pneu,xlim = c(2014,2020))
lines(test_pneu,col="red",lty=2)
```

# This data has Trending
differincing 1 step
```{r}
diff_train_pneu <- diff(train_pneu)
plot(diff_train_pneu)
```
# Augmented dickey fuller test

```{r}
library(tseries)
adf.test(diff_train_pneu,alternative = "stationary")
```
P-value<0.05 ,rej H0
that is the data  is stationary

# plot acf and pacf
```{r}
acf(as.vector(diff_train_pneu),lag.max = 60)
pacf(as.vector(diff_train_pneu),lag.max = 60)
```
(model 1) 
Seasonal effect 12 month = MA(1)
Non Seasonal effect = MA(2)
SMA(1)^12 , MA(2)

(model 2)
Seasonal effect 12 month = MA(1)
Non Seasonal effect = AR(2)
SMA(1)^12 , AR(2)

```{r}
model1 <- arima(diff_train_pneu,order=c(0,0,2),seasonal = list(order=c(0,0,1),period = 12))
model1
```
Check assumption
```{r}
library(TSA)
par(mfrow=c(1,2))
hist(rstandard(model1),xlab="Standardised residuals",main="")
qqnorm(rstandard(model1))
qqline(rstandard(model1))
tsdiag(model1,gof=30,omit.initial=T)
```

```{r}
model2 <- arima(diff_train_pneu,order=c(2,0,0),seasonal = list(order=c(0,0,1),period = 12))
model2
```
Check assumption
```{r}
library(TSA)
par(mfrow=c(1,2))
hist(rstandard(model2),xlab="Standardised residuals",main="")
qqnorm(rstandard(model2))
qqline(rstandard(model2))
tsdiag(model2,gof=30,omit.initial=T)
```
By diagnostic testing 
use MODEL1 -> SMA(1)^12 x MA(2) -> aic = 1619.62
```{r}
predict_diff_pneu <- predict(model1,23,prediction.interval = TRUE, level=0.95)
predict_pneu <- ts(train_pneu[108]+cumsum(predict_diff_pneu$pred),start=c(2018,1),frequency=12)
plot(test_pneu)
lines(predict_pneu,lty = 2,col ="red")
```
# try VAR model

```{r}
dis_dat <- data[c(2,3)]
# Change columns name
names(dis_dat) = c("Pneunomia","Pulmonary_TB")
# Time series dataframe
ts_data <- ts(dis_dat,start = c(2009,1),frequency = 12)
plot(ts_data)
```

### tseries - standard test adt.test ###
```{r}
library(tseries)
apply(ts_data, 2, adf.test)
```
both factors are Non stationary 
### diff 1 time
```{r}
diff_ts <- diff(ts_data)
apply(diff_ts, 2, adf.test)
```
both factors are Stationary 

### VAR
```{r}
plot.ts(diff_ts)
```
### Lag order identification ###
```{r}
library(vars)
VARselect(diff_ts, type = "none", lag.max = 10)
```
### Creating a VAR model with vars, VAR(2) and VAR(5) ###
VAR(2)

```{r}
var.model1 <- VAR(diff_ts, p=2, type="none")
summary(var.model1)
```
### Creating a VAR model with vars, VAR(2) and VAR(5) ###
VAR(5)

```{r}
var.model2 <- VAR(diff_ts, p=5, type="none")
summary(var.model2)
```
### Residual diagnostics ### Portmanteau test

```{r}
serial.test(var.model2)
```
reject H0 -> not independent

### Granger test for causality ###

```{r}
causality(var.model2, cause = c("Pneunomia"))
causality(var.model2, cause = c("Pulmonary_TB"))

#selecting variable have influence
#choose FTR
```
### 7. Forecasting VAR models

```{r}
fcast = predict(var.model2, n.ahead = 24)
plot(fcast)
```


```{r}
```

```{r}
```


```{r}
```


```{r}
```


```{r}
```

